Bibat: Batteries-include Bayesian Analysis Template

### 2024-01-29 paper/main.qmd
Update paper

---
title: "Bibat: Batteries-include Bayesian Analysis Template"
keywords: Bayesian inference,
abstract: >
  Despite their suitability for many scientific problems and the existence of
  sound theoretical and computational frameworks, adoption of Bayesian workflow
  in science remains a challenge. One reason for the difficulty is that it is
  difficult to write software that implements non-trivial Bayesian workflows.
  We aimed to address this difficulty by developing Bibat, a Python package
  providing  an interactive template that targets Bayesian statistical analysis
  projects. Bibat trivialises the otherwise painful process of structuring such
  a project and integrating together tools that help with individual parts of a
  Bayesian workflow, without compromising flexibility, scalability,
  interoperability or reproducibility. Bibat is available on the Python
  Package index,  documented at <https://bibat.readthedocs.io/> and
  developed at <https://github.com/teddygroves/bibat/>.
number-sections: true
author:
  - name: "Teddy Groves"
    orcid: 0000-0002-7109-3270
    email: tedgro@biosustain.dtu.dk
    affiliations:
      - id: biosustain
        name: DTU
        department: The Novo Nordisk Foundation Center for Biosustainability
        city: Kongens Lyngby
        country: Denmark
    attributes:
      corresponding: true
bibliography: bibliography.bib
format:
  html:
    embed-resources: true
  pdf:
    keep-tex: true
---

Many areas of science would benefit from replacing traditional statistical
methodologies with Bayesian workflow, as described in, for example
[@gelmanBayesianWorkflow2020;@grinsztajnBayesianWorkflowDisease2021;
@gabryVisualizationBayesianWorkflow2019].

There are now software tools that address most individual aspects of a Bayesian
workflow, from fetching and manipulating data to specifying, computing, storing,
inspecting and documenting statistical inferences. However, writing a software
project implementing a Bayesian workflow remains challenging due partly to the
difficulty of orchestrating and configuring these elements.

Bibat is a Python package that addresses this difficulty by providing an
interactive template for Bayesian workflow projects.

# Installation and usage

Bibat is installed by running the command `pip install bibat` and then used by
running the command `bibat`.

This command triggers an interactive form which prompts the user for
configuration information including project and repository name, author name, a
short description and choices of open source licence options, documentation
formats and whether to include tests and continuous integration.

Bibat then creates a folder with the chosen repository name in the current
working directory, containing code that implements bibat's example analysis.

# Documentation

Bibat is documented at <https://github.com/teddygroves/bibat/>. The
documentation website includes instructions for getting started, a detailed
explanation of bibat's concepts, an extended vignette illustrating intended
usage, full description of the python API and command line interface,
instructions for contributing and a section discussing accessibility
considerations.

Bibat's documentation website is continuously checked to ensure that it
implements accessibility best practices.

# Software

Bibat is implemented using Python 3. Its main dependencies are cookiecutter
[@greenfeldCookiecutter2021], pydantic [@pydanticdevelopersPydantic2022] and
click [@clickdevelopersClickPythonComposable2022].

Bibat projects have the following dependencies:

* Python 3 [@vanrossumPythonReferenceManual2009] and standard scientific Python
packages for data manipulation
* Stan [@carpenterStanProbabilisticProgramming2017] for specifying statistical
models and performing inferences
* cmdstanpy [@standevelopmentteamCmdStanPy2022] for interfacing between Python
and Stan
* arviz [@kumarArviZUnifiedLibrary2019] for storing and analysing completed
inferences
* pydantic and pandera [@niels_bantilan-proc-scipy-2020] for validation
* make [@stallman1991gnu] for automation.
 * Sphinx [@georgbrandlandthesphinxteamSphinx2022] and Quarto
[@Allaire_Quarto_2022] for documentation.

Bibat is continuously tested to ensure that it works on the operating systems
Linux, macOS and Windows.

Bibat's continuous integration runs on windows and ubuntu and runs bibat's test
suite as well as an end-to-end functional test on all supported Python versions.

# How bibat addresses specific Bayesian workflow difficulties

This section describes some specific problems that often affect Bayesian
workflow projects and which bibat helps to ameliorate.

## Data modelling

Any statistical analysis requires a definition of a prepared dataset, whether
implicitly or explicitly. Ideally the definition is explicit and allows for
arbitrarily many prepared datasets, accommodates both tabular and non-tabular
formats and provides functionality like validation and serialisation.

Bibat achieves this goal by providing abstract models for prepared data,
preconfigured to work well together, integrated into a data preparation pipeline
and able to be easily customised to suit any analysis.

Without a template such as bibat, users must either devise a new data modelling
framework from scratch as part of an analysis, which is time-consuming, or
else to miss out on either the flexibility or features that bibat's approach
provides.

## Reproducibility

Bibat ensures reproducibility by providing a preconfigured makefile with a
target `analysis` triggering creation of an isolated environment, installation
of dependencies, data preparation, statistical computation and analysis of
results. In this way a bibat analysis can be reproduced on most platforms
using a single command.

@@

## Collaboration

Bibat provides a preconfigured test environment, continuous integration, linting
and pre-commit hooks, making it suitable for collaborative software development.
In addition, including documentation as a first class component of the analysis
addresses a common problem in academic statistics projects where the paper gets
out of sync with the code.

## Installing dependencies

@@

@@

As part of a Bayesian workflow it is often necessary to fit a model and dataset
in different ways. For example, one might perform MCMC sampling of both the
prior and posterior distributions, perform multiple leave-out-one-fold fits for
cross-validation or need to compare MCMC sampling with an optimisation-based
alternative.

Bibat accommodates this ubiquitous scenario by introducing an abstraction called
"fitting mode". This abstraction allows bibat projects to handle fitting a
model and dataset in different ways appropriately and flexibly. For example, the
provided prior sampling fitting mode creates a Stan input dictionary with the
`likelihood` data variable set to `0`, performs MCMC sampling and writes data
to the `InferenceData` group `prior`. Bibat provides fitting modes corresponding
for prior sampling, posterior sampling and kfold posterior sampling. Users can
easily add additional fitting modes by copying these examples.

## Configure fitting concisely.

Bibat provides a Pydantic class `InferenceConfiguration`, parsed from toml files
in the `inferences` directory. By writing a toml file per inference the user can
configure statistical model, dataset, fitting modes and computational settings
(globally or per fitting mode).

This approach appropriately separates configuration from logic and avoids
unnecessary duplication of configuration for runs with the same model and
dataset.

## Storing MCMC samples

Bibat provides code that automatically saves its output `InferenceData` objects
in zarr format. This format distributes information from the same set of samples
into multiple smaller files, thereby helping to avoid file size limits of online
repository hosting services.

@@

There is some software that addresses the general task of facilitating Bayesian
workflow, but differently from bibat. For example, bambi [@capretto2020]
and brms [@burknerBrmsPackageBayesian2017] aim to make implementing Bayesian
workflows easier by providing ergonomic ways to specify and fit Bayesian
generalised models to tabular datasets. Bibat is complementary with these
packages, as it targets use cases that they do not support, such as analyses
where complex datasets or models might be required.

#

>>>>

---
title: "Bibat: Batteries-include Bayesian Analysis Template"
abstract: >
  Bayesian statistical workflow offers a powerful way to learn from data,
  but software software projects that implement complex Bayesian workflows
  in practice are unusual, partly due to the difficulty of orchestrating
  Bayesian statistical software. Bibat addresses this challenge by providing a
  full-featured, scalable Bayesian statistical analysis project
  using an interactive template. Bibat is available on the Python Package index,
  documented at <https://bibat.readthedocs.io/> and developed at
  <https://github.com/teddygroves/bibat/>. This paper explains the motivation
  for bibat, briefly describes intended usage, discusses key design choices,
  and reviews several examples of bibat's use in scientific applications.
number-sections: true
author:
  - name: "Teddy Groves"
    orcid: 0000-0002-7109-3270
    email: tedgro@biosustain.dtu.dk
    affiliations:
      - id: 1
        name: Danish Technical University
        department: The Novo Nordisk Foundation Center for Biosustainability
        city: Kongens Lyngby
        country: Denmark
    attributes:
      corresponding: true

bibliography: bibliography.bib

format:
  html:
    embed-resources: true
  acm-pdf:
    keep-tex: true

acm-metadata:
  # comment this out to make submission anonymous
  # anonymous: true

  # comment this out to build a draft version
  final: true

  # comment this out to specify detailed document options
  acmart-options: sigconf, review

  # acm preamble information
  copyright-year: 2024
  acm-year: 2024
  copyright: acmcopyright
  # doi: XXXXXXX.XXXXXXX
  conference-acronym: "KDD"
  conference-name: "KDD"
  conference-date: Aug 25--29, 2024
  conference-location: Barcelona, Esp
  keywords:
    - Python
    - Bayesian workflow
    - Methodology
  teaser:
    image: img/workflow.png
    caption: |
      Schematic representation of a Bayesian workflow implemented using bibat.
      The author inspects their analysis's results, edits files corresponding
      to the boxes on the left, runs the command `make analysis`, then repeats.
      Note that this workflow is modular, accommodates plurality and results in
      a final analysis that can be fully reproduced using a single command.
    description: "DESCRIPTION"

---

# Introduction: the problem of orchestrating Bayesian workflow software

The idea that Bayesian statistical analysis comprises not just
inference, but also specific approaches to related activities like data
preparation, model design, diagnosis, debugging and criticism, dates
at least to @boxBayesianInferenceStatistical1992. There is increasing
scholarly recognition of the need for a holistic view of "Bayesian
workflow" [@gelmanBayesianWorkflow2020;@grinsztajnBayesianWorkflowDisease2021;
@gabryVisualizationBayesianWorkflow2019]. Relatedly, software tools now
exist that address most individual aspects of a Bayesian workflow: see
@strumbeljPresentFutureSoftware for a review of the state of the art.

Unfortunately, currently available tools typically address one, or at most
a handful, of Bayesian workflow activities; it is left to the individual
project team to orchestrate the components. Writing software that performs
this orchestration can be time-consuming and tricky, especially in the common
scenario where it is not initially clear how many, or what kind of, statistical
models, datasets, data manipulations or investigations an analysis will require.

Bibat addresses this difficulty by providing a full-featured, high-quality
Bayesian workflow project that can be extended to implement a wide range of
statistical analyses.

# How bibat works

## Installation and usage

Bibat can be installed on the Windows, linux or macos command line on by running
the command `pip install bibat` and then used by running the command `bibat`.
This command triggers an interactive form which prompts the user to select a
range of customisation options. Bibat then creates a new directory containing
code that implements an example analysis, with customisations reflecting the
user's choices. This analysis works immediately, and can be reproduced with
the single command `make analysis` without the need for any further action by
the user: in this sense bibat comes with "batteries included".

## Documentation

Bibat is documented at <https://github.com/teddygroves/bibat/>. The
documentation website includes instructions for getting started, a detailed
explanation of bibat's concepts and an extended vignette illustrating how to
implement a complex statistical analysis starting from bibat's example analysis
usage. In addition, the documentation site contains a full description of
bibat's python API and command line interface, instructions for contributing and
a section discussing accessibility considerations.

# Design choices

Bibat's design was informed by these considerations.

## Accommodate a wide range of statistical analyses

Bibat projects explicitly allow for plurality at the level of input datasets,
prepared datasets, statistical models, fitting modes, computation methods and
analyses. In addition, bibat ensures that there are minimal restrictions on the
kind of components: for example, datasets need not be singular or tabular, and
statistical models need not be representable in formula syntax.

Thanks to these accommodations a project team using bibat should typically not
need to foresee the ultimate requirements of their analysis before starting
the project.

## Encourage reproducibility

Bibat provides a preconfigured makefile with a target `analysis` triggering
creation of an isolated environment, installation of dependencies, data
preparation, statistical computation and analysis of results. In this way a
bibat analysis can be reproduced on most platforms using a single command.

@@

## Use widely-adopted, open source and actively developed tools

Bibat projects are is written in modern Python and uses pydantic
@pydanticdevelopersPydantic2022 and pandera [@niels_bantilan-proc-scipy-2020]
for data modelling, Stan [@carpenterStanProbabilisticProgramming2017] for
statistical inference, cmdstanpy [@standevelopmentteamCmdStanPy2022] for
python-Stan interface, arviz [@kumarArviZUnifiedLibrary2019] for storing and
analyzing inferences and sphinx [@georgbrandlandthesphinxteamSphinx2022] and
quarto [@Allaire_Quarto_2022] for documentation.

Bibat itself is also written in modern Python and uses the popular
tools cookiecutter [@greenfeldCookiecutter2021], pydantic and click
[@clickdevelopersClickPythonComposable2022].

@@

## Implement community standards and best practices for collaborative software development

Bibat projects include a preconfigured test environment, continuous integration,
linting and pre-commit hooks, making them suitable for collaborative software
development. In addition, including documentation as a first class component of
the analysis addresses a common problem in academic statistics projects where
the paper gets out of sync with the code.

Bibat itself is continuously tested to ensure that it works on the operating
systems Linux, macOS and Windows. Bibat's continuous integration runs a test
suite as well as an end-to-end functional test on all supported Python versions.

Bibat is part of the PyOpenSci ecosystem, allowing for community help with
maintenance as well as peer review for code and documentation quality, usability
and accessibility. The PyOpenSci peer review for bibat can be found here: <https://github.com/pyOpenSci/software-submission/issues/83>

# Challenges

This section describes some specific challenges that often affect Bayesian
workflow projects and how bibat addresses them.

## Address complexity using a modular, file based approach

As discussed in @gelmanBayesianWorkflow2020, Bayesian workflows are complicated,
featuring plurality, cyclicity and complexity at many levels. Bibat accommodates
this by separating non-interacting analysis components into modules and by
serialising data to files wherever possible. Prepared datasets, statistical
models, inference configurations, inference results, plots and analyses all
have file representations. Fitting modes, data manipulations and data models are
modularised in code through the use of appropriately structured data classes
and functions.

Thanks to this approach it is possible to perform small sub analyses
individually and to iteratively add components without needing to consider
everything at once.

@@

As part of a Bayesian workflow it can be necessary to fit a model and dataset
in different ways. For example, one might perform MCMC sampling of both the
prior and posterior distributions, perform multiple leave-out-one-fold fits for
cross-validation or need to compare MCMC sampling with an optimisation-based
alternative.

Bibat accommodates this need by introducing an abstraction called "fitting
mode". This abstraction allows bibat projects to handle fitting a model
and dataset in different ways appropriately and flexibly. For example, the
provided prior sampling fitting mode creates a Stan input dictionary with the
`likelihood` data variable set to `0`, performs MCMC sampling and writes data
to the `InferenceData` group `prior`. Bibat provides fitting modes corresponding
for prior sampling, posterior sampling and k-fold posterior sampling. Users can
easily add additional fitting modes by copying these examples.

# Post launch use

Bibat has been used for several real statistical analyses.

@grovesteddyDgfreg2023 used bibat to compare a Bayesian and two non-Bayesian
approaches to modelling a biochemical thermodynamics dataset. Bibat facilitated
this analysis even though it was not very large by providing the fitting mode
abstraction, which was very useful for comparing the different methods.

In @grovesBaseball2022, Bibat was used to implement a sports analysis involving
two datasets, two models and four inferences, demonstrating that the generalised
Pareto distribution can be used to describe hitting ability in baseball. This
analysis is now included in bibat as an illustration, along with an accompanying
tutorial.

:::{#fig-baseball}

![](img/baseball.png){width=50%}

A graphical posterior predictive check produced as part of a bibat
analysis that fit two statistical models to two datasets of baseball data.
The coloured lines show each model's posterior predictive distributions
and the black dots show the two observed datasets.  See <https://github.com/teddygroves/bibat/tree/main/bibat/examples/baseball>
for the full analysis.

:::

In @grovesteddySphincter2024, bibat was used to implement an analysis of
cerebrovascular data from mice, involving two raw datasets, 6 prepared
datasets, 15 models and 15 inferences. In this case bibat's modular design made
it straightforward to iteratively add data transformations and models while
maintaining reproducibility.

These cases illustrate that bibat can be useful in a variety of real Bayesian
workflows, with different sizes, subject matters and emphases.

Bibat has a growing user community, with 16 GitHub stars at the time of writing.

@@

There is some software that addresses the general task of facilitating
Bayesian workflow, but using a different approach from bibat's. For example,
bambi [@capretto2020] and brms [@burknerBrmsPackageBayesian2017] aim to make
implementing Bayesian workflows easier by providing ergonomic ways to specify
and fit Bayesian regression models to tabular datasets. Bibat is complementary
with these packages, as it targets use cases that they do not support, such as
analyses where complex datasets or custom models might be required.

# Statements and Declarations

The authors have no financial or non-financial interests that are directly or
indirectly related to the work submitted for publication.
